---
title: "Analysis_Template"
author: "Alice Hong"
date: "2026-04-27"
output: pdf_document
---

# Libraries
```{r}
library(readxl)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(rstatix)
library(ggpubr)
```

# Load Data
- Replace file paths here with your directory path!
```{r}
# Load in metadata
# Example:
# meta <- read_excel('~/projects/cellpose-biosensor/data/P015-AH004.xlsx', sheet = 'Staining_Imaging')

meta <- read_excel('*REPLACE THIS*', sheet = 'Staining_Imaging')

# Load in counts
# Running the pipeline will have created a 'filter_steps.csv'
log <- read.csv('*REPLACE THIS WITH DIR PATH*/data/results/filter_steps.csv') %>%
  separate(stem, into = c("Imaged_On", "Exp", "Region", "replicate"), sep = "[-_]")

# Load all csv's and combine into 1 df
results_dir <- '*REPLACE THIS WITH DIR PATH*/data/results'
csv_files <- list.files(results_dir, pattern = '\\.csv$', full.names = TRUE)
csv_files <- csv_files[!grepl('filter_steps', csv_files)]
csv_files <- csv_files[sapply(csv_files, function(f) length(readLines(f)) > 1)]
```

# Run the chunk below once, then comment out. It will create a large .csv file with all data
```{r}
# Big df with all data + meta data!
# cells <- lapply(csv_files, read.csv) %>% bind_rows() %>%
#   separate(image_id, into = c("Imaged_On", "Exp", "Region", "replicate"), sep = "[-_]", remove = FALSE) %>%
#   left_join(meta %>% select(`Exp #`, Sex, `Experimental Group`), by = c("Exp" = "Exp #"))
```

# Uncomment below to filter out low count images...
# For example, in mAH004, removed C1M2 - dim BFP flagged during imaging
# Replace the file back for the location 
```{r}
# # Uncomment below to filter out low count images too..,
# keep_images <- log %>%
#   filter(n_size_filter >= 100) %>%
#   mutate(image_id = paste0(Imaged_On, "-", Exp, "_", Region, "_", replicate)) %>%
#   pull(image_id)
# 
# cells <- cells %>% filter(image_id %in% keep_images, Exp != "C1M2")

# cells <- cells %>%filter(Exp != "C1M2")
# 
# cat("No. NA values in cells df:", sum(is.na(cells)), "\n")

#write.csv(cells, file = "~/projects/cellpose-biosensor/data/results/cells.csv")

cells <- read.csv("~*REPLACE WITH FILE PATH*/data/results/cells.csv", row.names = 1)
```

# Check distribution of cell counts per image
```{r}
# Plot distribution of counts
log %>%
  mutate(stem = paste(Imaged_On, Exp, Region, replicate, sep = "-")) %>%
  ggplot(aes(x = reorder(stem, n_size_filter), y = n_size_filter)) +
  geom_col(fill = "#486dad") +
  coord_flip() +
  labs(x = NULL, y = "Cell count (after size filter)", title = "Cells per image") +
  theme_bw()
```
# Check distribution of BFP per image
```{r}
# cells %>%
#   ggplot(aes(x = reorder(image, bfp, FUN = median), y = bfp)) +
#   geom_boxplot(outlier.shape = NA) +
#   coord_flip() +
#   scale_y_continuous(trans = "log1p") +
#   theme_bw() +
#   labs(title = "BFP intensity per image", x = "Image (ordered by median BFP)", y = "BFP") +
#   theme(panel.grid = element_blank())
```


# Check which mice/regions the data comes from
```{r}
cells <- cells %>%
  mutate(image = paste(Imaged_On, Exp, Region, replicate, sep = "_"))

cell_summary <- cells %>%
  group_by(`Experimental.Group`, Sex, Region) %>%
  summarise(
    n_cells = n(),
    n_images = n_distinct(image),
    n_mice = n_distinct(Exp),
    .groups = "drop"
  )

cell_summary

image_summary <- cells %>%
  group_by(image, `Experimental.Group`, Sex, Region, Exp) %>%
  summarise(
    n_cells = n(),

    # RFP/BFP
    mean_rfp_bfp   = mean(rfp_over_bfp, na.rm = TRUE),
    median_rfp_bfp = median(rfp_over_bfp, na.rm = TRUE),
    sd_rfp_bfp     = sd(rfp_over_bfp, na.rm = TRUE),
    iqr_rfp_bfp    = IQR(rfp_over_bfp, na.rm = TRUE),
    q25_rfp_bfp    = quantile(rfp_over_bfp, 0.25, na.rm = TRUE),
    q75_rfp_bfp    = quantile(rfp_over_bfp, 0.75, na.rm = TRUE),

    # GFP/BFP
    mean_gfp_bfp   = mean(gfp_over_bfp, na.rm = TRUE),
    median_gfp_bfp = median(gfp_over_bfp, na.rm = TRUE),
    sd_gfp_bfp     = sd(gfp_over_bfp, na.rm = TRUE),
    iqr_gfp_bfp    = IQR(gfp_over_bfp, na.rm = TRUE),
    q25_gfp_bfp    = quantile(gfp_over_bfp, 0.25, na.rm = TRUE),
    q75_gfp_bfp    = quantile(gfp_over_bfp, 0.75, na.rm = TRUE),

    .groups = "drop"
  )
```

# Add later to chunk above (more statsss)
```{r}
# prop_high_rfp = mean(rfp_over_bfp > 0.5, na.rm = TRUE),
# prop_high_gfp = mean(gfp_over_bfp > 0.5, na.rm = TRUE),
# cv_rfp_bfp = sd_rfp_bfp / mean_rfp_bfp,
# cv_gfp_bfp = sd_gfp_bfp / mean_gfp_bfp,
```

# Quick GROUPED boxplot of rfp/bfp and gfp/bfg colroed by sex and facet by region, grouped by condition
# Plotted based off raw data, not taking into account differences in N of cells per image
```{r}
cells %>%
  ggplot(aes(x = `Experimental.Group`, y = rfp_over_bfp, fill = Sex)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~Region) +
  scale_y_continuous(transform="log1p", limits=c(0,0.75)) +
  #scale_y_continuous(limits=c(0,1)) +
  theme_bw() +
  labs(x = NULL, y = "RFP / BFP") +
  theme(panel.grid = element_blank())

cells %>%
  ggplot(aes(x = `Experimental.Group`, y = gfp_over_bfp, fill = Sex)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~Region) +
  scale_y_continuous(transform="log1p", limits=c(0,0.75)) +
  #scale_y_continuous(limits=c(0,1)) +
  theme_bw() +
  labs(x = NULL, y = "GFP / BFP") +
  theme(panel.grid = element_blank())
```

# median signal per image, grouped by condition
```{r}
p_rfp <- image_summary %>%
  ggplot(aes(x = `Experimental.Group`, y = median_rfp_bfp, fill = Sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  #geom_jitter(aes(color = Sex), width = 0.15, size = 2, alpha = 0.5) +
  geom_jitter(aes(color = Sex), 
              position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), 
              size = 1.5,
              alpha = 0.4) +
  facet_wrap(~Region) +
  #scale_y_continuous(trans = "log1p") +
  labs(
    title = "RFP/BFP",
    x = NULL,
    y = "Median RFP/BFP"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank())

p_rfp

p_gfp <- image_summary %>%
  ggplot(aes(x = `Experimental.Group`, y = median_gfp_bfp, fill = Sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  #geom_jitter(aes(color = Sex), width = 0.15, size = 2, alpha = 0.8) +
  geom_jitter(aes(color = Sex), 
            position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), 
            size = 1.5,
            alpha = 0.4) +
  facet_wrap(~Region) +
  #scale_y_continuous(trans = "log1p") +
  labs(
    title = "GFP/BFP",
    x = NULL,
    y = "Median GFP / BFP"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank())

p_gfp
```


# median signal per image, grouped by condition only
```{r}
p_rfp <- image_summary %>%
  ggplot(aes(x = `Experimental.Group`, y = median_rfp_bfp, color = `Experimental.Group`)) +
  geom_violin(alpha = 0.6) +
  #geom_jitter(aes(color = Sex), width = 0.15, size = 2, alpha = 0.5) +
  geom_jitter(
              position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), 
              size = 1.5,
              alpha = 0.4) +
  #facet_wrap(~Region) +
  #scale_y_continuous(trans = "log1p") +
  labs(
    title = "RFP/BFP",
    x = NULL,
    y = "Median RFP/BFP"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank(), legend.position = "none") 

p_rfp


p_gfp <- image_summary %>%
  ggplot(aes(x = `Experimental.Group`, y = median_gfp_bfp, color = `Experimental.Group`)) +
  geom_violin(alpha = 0.6) +
  #geom_jitter(aes(color = Sex), width = 0.15, size = 2, alpha = 0.8) +
  geom_jitter(
            position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75), 
            size = 1.5,
            alpha = 0.4) +
  #facet_wrap(~Region) +
  #scale_y_continuous(trans = "log1p") +
  labs(
    title = "GFP/BFP",
    x = NULL,
    y = "Median GFP / BFP"
  ) +
  theme_bw() +
  theme(panel.grid = element_blank(), legend.position = "none")

p_gfp
```


# BFP Threshold?

## Overall BFP distribution
```{r}
cells %>%
  ggplot(aes(x = bfp)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 100,
    fill = "grey70",
    #color = "black",
    alpha = 0.8
  ) +
  geom_density(
    color = "black",
    linewidth = 1
  ) +
  theme_bw() +
  labs(
    title = "Distribution of BFP intensity across cells",
    x = "BFP intensity",
    y = "Density"
  ) +
  theme(panel.grid = element_blank())
```
## BFP distribution by experimental group
```{r}
cells %>%
  ggplot(aes(x = bfp, fill = Experimental.Group)) +
  geom_density(alpha = 0.4) +
  #scale_x_continuous(trans = "log1p") +
  theme_bw() +
  labs(
    title = "BFP intensity distribution by group",
    x = "BFP intensity",
    y = "Density"
  ) +
  theme(panel.grid = element_blank())
```
## BFP by group/region/sex
```{r}
cells %>%
  ggplot(aes(x = Experimental.Group, y = bfp, fill = Sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  facet_wrap(~ Region) +
  theme_bw() +
  labs(
    title = "BFP intensity by group and region",
    x = NULL,
    y = "BFP intensity"
  )

## Cutting off some outliers
cells %>%
  ggplot(aes(x = Experimental.Group, y = bfp, fill = Sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  facet_wrap(~ Region) +
  theme_bw() +
  scale_y_continuous(limits = c(0,5000)) +
  labs(
    title = "BFP intensity by group and region",
    x = NULL,
    y = "BFP intensity"
  )

bfp_summary <- cells %>%
  group_by(Experimental.Group, Sex, Region) %>%
  summarise(
    n_cells = n(),
    
    mean_bfp   = mean(bfp, na.rm = TRUE),
    median_bfp = median(bfp, na.rm = TRUE),
    
    sd_bfp  = sd(bfp, na.rm = TRUE),
    iqr_bfp = IQR(bfp, na.rm = TRUE),
    
    q25_bfp = quantile(bfp, 0.25, na.rm = TRUE),
    q75_bfp = quantile(bfp, 0.75, na.rm = TRUE),
    
    min_bfp = min(bfp, na.rm = TRUE),
    max_bfp = max(bfp, na.rm = TRUE),
    
    .groups = "drop"
  )

bfp_summary
```

# Remove below 800 BFP cells??
```{r}
# cells_filt <- cells %>% filter(bfp > 1000)
# 
# image_summary_filt <- cells_filt %>%
#   group_by(image, Experimental.Group, sex, region, exp) %>%
#   summarise(
#     n_cells = n(),
#     
#     median_rfp_bfp = median(rfp_over_bfp, na.rm = TRUE),
#     mean_rfp_bfp   = mean(rfp_over_bfp, na.rm = TRUE),
#     
#     median_gfp_bfp = median(gfp_over_bfp, na.rm = TRUE),
#     mean_gfp_bfp   = mean(gfp_over_bfp, na.rm = TRUE),
#     
#     .groups = "drop"
#   )

# p_rfp_filt <- image_summary_filt %>%
#   ggplot(aes(x = Experimental.Group, y = median_rfp_bfp, fill = sex)) +
#   geom_boxplot(
#     position = position_dodge(width = 0.75),
#     outlier.shape = NA,
#     alpha = 0.6
#   ) +
#   geom_jitter(
#     aes(color = sex),
#     position = position_jitterdodge(
#       jitter.width = 0.1,
#       dodge.width = 0.75
#     ),
#     size = 2,
#     alpha = 0.8
#   ) +
#   facet_wrap(~region) +
#   theme_bw() +
#   labs(
#     title = "RFP/BFP per image (BFP > 800)",
#     x = NULL,
#     y = "Median RFP / BFP"
#   )
# 
# p_rfp_filt
# p_rfp
```


# STATS??? 

Small-ish n (images)
Non-normal (skewed, bounded near 0)
Independent units = images
Grouped by Region + Sex

Wilcoxon rank-sum test (Mann–Whitney U)

## between groups (within region), between sexes (within group + region)
```{r}
# stats_group <- image_summary %>%
#   group_by(region) %>%
#   wilcox_test(median_rfp_bfp ~ Experimental.Group) %>%
#   adjust_pvalue(method = "BH") %>%
#   add_significance()
# 
# stats_group
# 
# stats_sex <- image_summary %>%
#   group_by(region, Experimental.Group) %>%
#   filter(n_distinct(sex) == 2) %>%   # keep only groups with both M and F
#   wilcox_test(median_rfp_bfp ~ sex) %>%
#   adjust_pvalue(method = "BH")
# 
# stats_sex
```

## Within each group + sex, do regions differ at all?

- Across regions (3), so need multi gorup test like ANOVA, but not parametric so kruskal wallis?

```{r}
# stats_region <- image_summary %>%
#   group_by(Experimental.Group, sex) %>%
#   kruskal_test(median_rfp_bfp ~ region)
# 
# stats_region
# 
# stats_region_pw <- image_summary %>%
#   group_by(Experimental.Group, sex) %>%
#   pairwise_wilcox_test(
#     median_rfp_bfp ~ region,
#     p.adjust.method = "BH"
#   )
# 
# stats_region_pw
```

# A different kind of plot - PDFS?

## RFP/BFP PDFs
```{r}
p_rfp_density_female <- cells %>%
  filter(Sex == "female") %>%
  ggplot(aes(x = rfp_over_bfp, color = Experimental.Group, fill = Experimental.Group)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~Region) +
  scale_x_continuous(trans = "log1p") +
  theme_bw() +
  labs(
    title = "RFP/BFP distribution across cells, female",
    x = "RFP / BFP (log1p scale)",
    y = "Density",
    color = "Group",
    fill = "Group"
  ) +
  theme(panel.grid=element_blank())

p_rfp_density_female

p_rfp_density_male <- cells %>%
  filter(Sex == "male") %>%
  ggplot(aes(x = rfp_over_bfp, color = Experimental.Group, fill = Experimental.Group)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~Region) +
  scale_x_continuous(trans = "log1p") +
  theme_bw() +
  labs(
    title = "RFP/BFP distribution across cells, male",
    x = "RFP / BFP (log1p scale)",
    y = "Density",
    color = "Group",
    fill = "Group"
  ) +
  theme(panel.grid=element_blank())

p_rfp_density_male
```
```{r}
cells %>%
  #ggplot(aes(x = rfp_over_bfp, fill = Experimental.Group)) +
  ggplot(aes(x = rfp_over_bfp, group = image_id, color = Experimental.Group, fill = Experimental.Group)) +
  
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  facet_grid(Sex ~ Region) +
  #scale_x_continuous(trans = "log1p") +
  scale_x_continuous(limits=c(0,1)) +
  theme_bw() +
  labs(
    title = "Cell-level RFP/BFP distributions",
    x = "RFP / BFP (log1p scale)",
    y = "Cell count"
  ) + theme(panel.grid = element_blank())
```
# Now for GFP
```{r}
p_gfp_density_female <- cells %>%
  filter(Sex == "female") %>%
  ggplot(aes(x = gfp_over_bfp, color = Experimental.Group, fill = Experimental.Group)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~Region) +
  scale_x_continuous(trans = "log1p") +
  theme_bw() +
  labs(
    title = "GFP/BFP distribution across cells - Female",
    x = "GFP / BFP (log1p scale)",
    y = "Density",
    color = "Group",
    fill = "Group"
  ) + theme(panel.grid=element_blank())

p_gfp_density_female

p_gfp_density_male <- cells %>%
  filter(Sex == "male") %>%
  ggplot(aes(x = gfp_over_bfp, color = Experimental.Group, fill = Experimental.Group)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  facet_wrap(~Region) +
  scale_x_continuous(trans = "log1p") +
  theme_bw() +
  labs(
    title = "GFP/BFP distribution across cells - Male",
    x = "GFP / BFP (log1p scale)",
    y = "Density",
    color = "Group",
    fill = "Group"
  ) + theme(panel.grid=element_blank())

p_gfp_density_male
```


```{r}
cells %>%
  ggplot(aes(x = gfp_over_bfp, fill = Experimental.Group)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  facet_grid(Sex ~ Region) +
  scale_x_continuous(trans = "log1p") +
  theme_bw() +
  labs(
    title = "Cell-level GFP/BFP distributions",
    x = "GFP / BFP (log1p scale)",
    y = "Cell count"
  ) +theme(panel.grid=element_blank())
```

# Clean up column and variable names

```{r}
# library(janitor)
# 
# cells <- cells %>% clean_names()
# image_summary <- image_summary %>% clean_names()
# 
# cells <- cells %>%
#   mutate(Experimental.Group = recode(Experimental.Group,
#     "PEG + DSS" = "PEG_DSS",
#     "Water + Water" = "Water"
#   ))
# 
# image_summary <- image_summary %>%
#   mutate(Experimental.Group = recode(Experimental.Group,
#     "PEG + DSS" = "PEG_DSS",
#     "Water + Water" = "Water"
#   ))
```


# FC Calculations?

```{r}
# fc_summary <- image_summary %>%
#   group_by(region, sex, experimental_group) %>%
#   summarise(
#     mean_rfp   = mean(median_rfp_bfp, na.rm = TRUE),
#     median_rfp = median(median_rfp_bfp, na.rm = TRUE),
#     mean_gfp   = mean(median_gfp_bfp, na.rm = TRUE),
#     median_gfp = median(median_gfp_bfp, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# fc <- fc_summary %>%
#   pivot_wider(
#     names_from = experimental_group,
#     values_from = c(mean_rfp, median_rfp, mean_gfp, median_gfp)
#   ) %>%
#   mutate(
#     fc_mean_rfp   = mean_rfp_PEG_DSS / mean_rfp_Water,
#     fc_median_rfp = median_rfp_PEG_DSS / median_rfp_Water,
#     fc_mean_gfp   = mean_gfp_PEG_DSS / mean_gfp_Water,
#     fc_median_gfp = median_gfp_PEG_DSS / median_gfp_Water
#   )
# 
# fc %>%
#   select(region, sex, fc_mean_rfp, fc_mean_gfp, fc_median_rfp, fc_median_gfp)
```

